# *Poppr*: an R Package For Genetic Analysis of Populations With Clonal, Partially Clonal, and/or Sexual Reproduction

```{r, results = "asis", echo = FALSE}
out <- "
\\singlespacing
\\begin{center}
"
cat(beaverdown::iflatex(out))
```

Zhian N. Kamvar, Javier F. Tabima, and Niklaus J. Grünwald


```{r, results = "asis", echo = FALSE}
out <- "
\\end{center}
\\vspace*{\\fill}
"
cat(beaverdown::iflatex(out))
```
Journal: **PeerJ**

PO Box 614, Corte Madera, CA 94976, USA    

Published 2014-03-04, Issue: **2**:e281, 
DOI: [10.7717/peerj.281](https://dx.doi.org/10.7717/peerj.281)

```{r, results = "asis", echo = FALSE}
out <- "
\\doublespacing
\\newpage
"
cat(beaverdown::iflatex(out))
```

## Abstract


Many microbial, fungal, or oomcyete populations violate assumptions for
population genetic analysis because these populations are clonal, admixed,
partially clonal, and/or sexual. Furthermore, few tools exist that are
specifically designed for analyzing data from clonal populations, making
analysis difficult and haphazard. We developed the R package poppr providing
unique tools for analysis of data from admixed, clonal, mixed, and/or sexual
populations. Currently, poppr can be used for dominant/codominant and
haploid/diploid genetic data. Data can be imported from several formats
including GenAlEx formatted text files and can be analyzed on a user-defined
hierarchy that includes unlimited levels of subpopulation structure and clone
censoring. New functions include calculation of Bruvo's distance for
microsatellites, batch-analysis of the index of association with several indices
of genotypic diversity, and graphing including dendrograms with bootstrap
support and minimum spanning networks. While functions for genotypic diversity
and clone censoring are specific for clonal populations, several functions found
in poppr are also valuable to analysis of any populations. A manual with
documentation and examples is provided. Poppr is open source and major releases
are available on CRAN: http://cran.r-project.org/package=poppr. More supporting
documentation and tutorials can be found under 'resources' at:
http://grunwaldlab.cgrb.oregonstate.edu/.

## Introduction 


The Wright-Fisher model of populations is one of the oldest models
utilized in population genetic theory. Populations in this model are
characterized as having non-overlapping generations with a constant size
free from any selective pressures [@Weir_1997; @Charlesworth_1998;
@nielsen2013introduction]. Conceptually, these
populations are represented as pools of alleles that are independently
assorting where random mating is approximated by randomly sampling
alleles with replacement from one generation to the next. Assumptions of
this model, or related models, are implicitly assumed for common
population genetic analysis tools. In clonal populations, however,
alleles are not independently passed on from one generation to the next,
and these assumptions are violated. Classical textbooks on population
genetics do not provide much guidance on how to analyze clonal or mixed
clonal and sexual populations. In reality, many populations are not
strictly clonal or sexual, but can range from completely sexual to
completely clonal and this is commonly observed for fungal, oomycete, or
microbial populations [@anderson1995clonality; @milgroom1996recombination]. Currently, analysis of these populations
is not straightforward as we lack the sophisticated tools and methods
developed for model populations that are typically either haploid or
diploid [@grunwald2011evolution].

Inferring population structure with many commonly used model-based
clustering approaches such as the program 
<span style="font-variant:small-caps;">Structure</span>
[@pritchard2000inference] is inherently problematic for clonal
populations. These approaches cannot be used as clonal populations
violate basic assumptions of panmixia and Hardy-Weinberg equilibrium.
Thus, model free methods such as those relying on k-means clustering,
dendrograms including bootstrap support for clades, or minimum spanning
networks are more appropriate [@goss2009population; @cooke2012genome; @mascheretti2008reconstruction]. Furthermore,
analysis of mixed or clonal populations traditionally relies on
calculation of diversity of genotypes observed and analysis of
clone-censored versus non-censored populations [@mcdonald1997population;
@milgroom1996recombination; @Grunwald:2006]. Clone censoring involves
reduction of any population sample to a single observation for each
multilocus genotype (MLG) in a population thereby approximating
panmictic populations and removing the effect of genetic linkage
[@milgroom1996recombination]. Analysis of diversity, in turn, involves
calculation of the number of genotypes observed (richness), diversity,
and evenness [@Grunwald:2003]. Typical measures of genotypic diversity are
borrowed from ecology and use either the Shannon-Wiener or Stoddart and
Taylor index [@stoddart1988genotypic; @shannon2001mathematical; @Grunwald:2003].

A critical aspect of analyzing clonal or mixed populations is testing a
null hypothesis of panmixia [@milgroom1996recombination]. Testing of this
hypothesis for potentially clonal populations typically relies on
assessment of linkage disequilibrium among loci
[@milgroom1996recombination]. This is achieved via calculation of the
index of association or related indices in combination with resampling
of the data to obtain a null distribution for the expectation of random
mating [@burt1996molecular; @Brown:1980; @Smith:1993; @milgroom1996recombination]. These approaches
have, for example, been applied to *Pyrenophora teres*
[@peever1994genetic] and *Aphanomyces euteiches* [@Grunwald:2006] and are
routinely used in the analyses of clonal populations although they are
not easily calculated given available software including <span
style="font-variant:small-caps;">multilocus</span>, which is no longer
supported, and <span style="font-variant:small-caps;">LIAN</span>, which
only works for haploids [@Agapow:2001; @Haubold:2000].

Hierarchical sampling adds another layer of complexity to analysis of
clonal populations. With microbial populations, the geographic structure
of each population is not entirely clear, and it is often important to
sample temporally to see if clones persist over time [@Grunwald:2006]. A
common approach when faced with multiple levels of sampling is to create
a separate data set for each level or combination of levels and to
analyze them separately. However, the number of data sets undergo a
factorial increase with each hierarchical level, therefore increasing
the chances of human error in data reformatting or analysis. Thus, tools
are needed for analysis of population data across hierarchies or subsets
of data.

Here, we introduce the R package *poppr* that is specifically designed
for analysis of populations that are clonal, admixed and/or sexual.
*Poppr* complements and builds on previously existing R packages
including *adegenet* and *vegan* [@Jombart:2008; @Jombart:2011; @vegan]
while implementing tools novel to R significantly facilitating data
import, population genetic analyses, and graphing of clonal or partially
clonal populations. These tools include among others: analysis across
hierarchies of populations, subsetting of populations, clone-censoring,
Bruvo's genetic distance @Bruvo:2004, the index of association and
related statistics [@Brown:1980; @Smith:1993], and bootstrap support for
trees based on Bruvo's distance. By providing a centralized suite of
tools appropriate for many data types, this package represents a novel
and useful resource specifically tailored for analysis of clonal
populations.

## Materials and Methods


### Data import


*Poppr* allows import of data in several formats for
dominant/codominant, haploid/diploid and geographic data. The R package
*adegenet*, that defines the `genind` data structure that *poppr*
utilizes, allows support for importing data natively from <span
style="font-variant:small-caps;">Structure, Genetix, Genepop</span>, and
<span style="font-variant:small-caps;">Fstat</span>. While these formats
are very common and widely supported, these do not allow for import of
geographic and/or regional data. Furthermore, *adegenet* will only
handle diploids with this format, though manual import is possible. To
aid in importing data, *poppr* has newly added the function
`read.genalex()`, to read data from *GenAlEx* formatted text files into
the `genind` data object of the package *adegenet* [@Jombart:2008;
@Jombart:2011; @Peakall:2006]. *GenAlEx* is a popular add-in for <span
style="font-variant:small-caps;">Microsoft Excel</span> that can handle
data including codominant/dominant and haploid/diploid markers as well
as geographic and regional data. This function further facilitates the
import of haploid, geographic, and regional data.

Transferring data to new formats and manipulating data by hand, such as
collapsing data into clones or subsetting data into different
hierarchical levels, is tedious, creates redundancy, and can result in
lost or misrepresented data. *Poppr* includes tools to automate such
repetitive tasks. Many currently available data formats and software
implementations allow analysis of only one or two levels of a population
hierarchy. With *poppr* the user can import a single data set with an
unlimited number of hierarchical levels. This is achieved by having the
user combine the levels using a common delimiter (e.g.
"Year\_Country\_City"). These combined levels are then used as the
defining population factor in the input file and can easily be
manipulated within R.

### Data analysis


Once data is imported into R, the user can dynamically access and
manipulate the population hierarchy with the function `splitcombine()`,
subset the data set by population with `popsub()`, and check for cloned
multilocus genotypes using `mlg()`. For data sets that include clones,
the *poppr* function `clonecorrect()` will censor exact clones with
respect to any level of a population hierarchy by creating a new data
set that includes only unique multilocus genotypes (MLGs) per
population. A full list of functions available in *poppr* is provided in
table \@ref(tab:poppr1).


```{r, poptab1, results = 'asis', echo = FALSE}
tab <- if (LATEX) "data/poppr-table1.tex" else "data/poppr-table1.md"
out <- readLines(tab)
cat(out, sep = "\n")
```

Typical analyses in *poppr* start with summary statistics for diversity,
rarefaction, evenness, MLG counts, and calculation of distance measures
such as Bruvo's distance, providing a suitable stepwise mutation model
appropriate for microsatellite markers [@Bruvo:2004]. *Poppr* will define
MLGs in your data set, show where they cross populations, and can
produce graphs and tables of MLGs by population that can be used for
further analysis with the R package *vegan* @vegan. Many of the
diversity indices calculated by the *vegan* function `diversity()` are
useful in analyzing the diversity of partially clonal populations. For
this reason, *poppr* features a quick summary table (Table \[table2\])
that incorporates these indices along with the index of association,
$I_A$ [@Brown:1980; @Smith:1993], and its standardized form, $\bar{r}_d$,
which accounts for the number of loci sampled [@Agapow:2001]. Both
measures of association can detect signatures of multilocus linkage and
values significantly departing from the null model of no linkage among
markers are detected via permutation analysis utilizing one of four
algorithms described in table \[table3\] [@Agapow:2001]. The user can
specify the number of samples taken from the observed data set to obtain
the null distribution expected for a randomly mating population.
Detailed examples of these analyses can be found in the *poppr* manual.


```{r, poptab2, results = 'asis', echo = FALSE}
tab <- if (LATEX) "data/poppr-table2.tex" else "data/poppr-table2.md"
out <- readLines(tab)
if (LATEX){
	cap <- paste(readLines("data/poppr-table2caption.md"), collapse = "\n")
  cap <- beaverdown::render_caption(cap, figname = "popprtable2", to = TO)
  subcap <- "Summary table shown as it would appear in the R console produced by the \\texttt{poppr()} function with 999 permutations to calculate \\(I_A\\) and \\(\\bar{r}_d\\) p-values from the \\texttt{Aeut} data set in \\textit{poppr} from (Grünwald et al., \\protect\\hyperlink{ref-Grunwald:2003}{2003})."
  out[grep("caption", out)] <- paste0("\\caption[", subcap, "]{", cap, "}")
}
cat(out, collapse = "\n")
```

```{r, poptab3, results = 'asis', echo = FALSE}
tab <- if (LATEX) "data/poppr-table3.tex" else "data/poppr-table3.md"
out <- readLines(tab)
cat(out, sep = "\n")
```


### Visualizations


*Poppr* generates bar charts of MLG counts found within each population
of your data set (Fig. \[fig:FIGURE\_1\]). Histograms with rug plots for
$I_A$ and $\bar{r}_d$ allow visual assessment of the quality of the
distribution derived from resampling to see if a higher number of
replications are necessary (Fig. \[fig:FIGURE\_2\]). *Poppr*
automatically produces custom minimum spanning networks for Bruvo's or
other distances using Prim's algorithm, as implemented in the package
*igraph* @igraph, with the functions `bruvo.msn()` for Bruvo's distance
(Fig. \[fig:FIGURE\_3\]) and `poppr.msn()` for any distance matrix. The
combination of data structures from *adegenet* and *igraph* allow
graphing that is color coded by population with vertices grouped by MLG
[@Jombart:2008; @Jombart:2011; @igraph]. *Poppr* also includes
visualization of dendrograms using UPGMA @phangorn and Neighbor-Joining
@ape algorithms with bootstrap support for Bruvo's distance using the
function `bruvo.boot()` (Fig. \[fig:FIGURE\_4\]). Neither graphing of
minimum spanning networks or dendrograms with bootstrap support are
currently possible for populations in any other R packages.

### Performance


Most of the functions in *Poppr* were written and optimized for
performance in R and are available for inspection and/or download at
<https://github.com/grunwaldlab/poppr>. Algorithms of $\geq O(n^2)$
complexity were written in the byte-compiled C language to optimize
runtime performance.

For comparisons of $I_A$ and Bruvo's distance, we utilized the data set
'nancycats' (237 diploid individuals genotyped at nine microsatellite
loci) from the *adegenet* package. Calculations were run independently
10 times and then averaged. Bruvo's distance was calculated on a machine
with OSX 10.8.4 and a 2.9 GHz intel processor. The $I_A$ and $\bar{r}_d$
calculations were performed on a machine with OSX 10.5.8 and a 2.4 GHz
intel processor due to the inability of the software <span
style="font-variant:small-caps;">multilocus</span> to work on any later
version of OSX.

Citation of methods implemented in *poppr*.
-------------------------------------------

Several of the methods implemented in *poppr* are described elsewhere.
Users should refer to the original publications for interpretations and
citation. See table \[table4\] for a full list of citations. As with any
R package, users should always cite the R Core Team [@R]. 
Results and Discussion {#results-and-discussion .unnumbered}
======================

*Poppr* provides significant, convenient tools for analysis of clonal
and partially clonal populations available in one environment on all
major operating systems. The ability to analyze data for multiple
populations across a user-defined hierarchy and clone-censoring provide
novel functionality in R. Combined with R's graphing abilities,
publication-ready figures are thus obtained conveniently.

### New functionalities


*Poppr* implements several new functionalities. As of this writing,
aside from *poppr*, there exist two programs that calculate $I_A$: <span
style="font-variant:small-caps;">LIAN</span> [@Haubold:2000] and <span
style="font-variant:small-caps;">multilocus</span> [@Agapow:2001]. <span
style="font-variant:small-caps;">LIAN</span> can calculate $I_A$ for
haploid data and is only available online or for \*nix systems with a C
compiler such as OSX and Linux [@Haubold:2000]. <span
style="font-variant:small-caps;">Multilocus</span> implemented
$\bar{r}_d$, a novel correction for $I_A$, but is no longer supported
[@Agapow:2001]. <span style="font-variant:small-caps;">Multilocus</span>
will only calculate index values for one data set at a time and <span
style="font-variant:small-caps;">LIAN</span> requires the user to
structure the data set with populations in contiguous blocks to analyze
multiple populations within a single file. Thus *poppr* provides
significant improvements for calculation of linkage disequilibrium, and
handles both haploid and diploid data, works on all major operating
systems, and is capable of batch analysis of multiple files and multiple
populations defined within a file including the possibility of clone
correction and sub-setting. A comparison of the capabilities of these
programs are summarized in Table \[table5\].

To test significance for $I_A$ and $\bar{r}_d$, *poppr* offers four
permutation algorithms. Each one will randomly shuffle data at each
locus, effectively unlinking the loci. The algorithm previously utilized
by <span style="font-variant:small-caps;">mutlilocus</span> is included.
The <span style="font-variant:small-caps;">multilocus</span>-style
algorithm shuffles genotypes, maintaining the associations between
alleles at each locus [@Agapow:2001]. More appropriately, alleles are
expected to assort independently in panmictic populations. *Poppr* thus
provides three new algorithms for permutation that allow for independent
allele assortment at each locus. The default algorithm permutes the
alleles at each locus and the remaining two will randomly sample alleles
from a multinomial distribution parametrically and non-parametrically
[@Weir_1997]. Details of these algorithms are presented in table
\[table3\]. Because the index of association is calculated using a
binary measure of dissimilarity, we have also made this available as a
distance measure called `diss.dist()`. This pairwise distance is based
on the percent allelic differences.

*Poppr* also newly implements Bruvo's genetic distance that utilizes a
stepwise mutation model appropriate for microsatellite data [@Bruvo:2004]. While this distance is implemented in the program <span
style="font-variant:small-caps;">GenoDive</span> @genodive and the R
package *polysat* @polysat, there are a few caveats with these two
implementations. <span style="font-variant:small-caps;">GenoDive</span>
is closed-source, and only implemented in OSX. Both *poppr* and
*polysat* are open-source and available on all platforms, but *polysat*,
being optimized for polyploid individuals with ambiguous allelic dosage,
is inappropriate for analyzing diploids. *Polysat* will collapse
homozygous individuals into a single allele and attempt to infer the
second allelic state in comparison with heterozygous individuals. Since
haploid and diploid individuals show clear allelic dosage, this
procedure creates a bias misrepresenting the true distance. Not only is
*poppr* not subject to this bias, but it also newly introduces bootstrap
support for this distance as shown in Figure \[fig:FIGURE\_4\].

### Performance


*Poppr* reduces the amount of intermediate files and repetitive tasks
needed for basic population genetic analyses and implements
computationally intensive functions, such as Bruvo's distance and the
index of association in C to improve performance. The *polysat* package
calculation of Bruvo's distance took 58.3 seconds on average whereas
*poppr*'s calculation was over 190 times faster, averaging 0.3 seconds
(Table \[table6\]). For calculation of $I_A$ and $\bar{r}_d$ with 100
permutations and Nei's genotypic diversity [@Nei:1978], <span
style="font-variant:small-caps;">multilocus</span> required around 9.12
minutes on average, as compared to 13.4 seconds with *poppr*.

## Conclusions


The R package *poppr* provides new functions and tools specifically
tailored for analysis of data from clonal or partially clonal
populations. No software currently available provides this set of tools.
Novel capabilities include analysis across multiple populations at
multiple levels of hierarchies, clone-censoring, and subsetting. These
in combination with R's command line interface and scripting
capabilities makes analyses of these populations more streamlined and
tractable. By implementing computationally expensive algorithms such as
Bruvo's distance and $I_A$ in C, analyses of multiple populations that
would normally take hours to complete can now be finished in a matter of
minutes. This allowed us to expand the utility of these measures to
convenient new graphing abilities such as automatically creating
dendrograms with bootstrap support for Bruvo's distance and minimum
spanning networks. While major releases of *poppr* are available on
CRAN, we are continuing to develop this package to be able to
efficiently handle genome-sized SNP data. Development versions are
available on github at <https://github.com/grunwaldlab/poppr>.

## Acknowledgements


The authors would like to thank Sydney Everhart and Corine Schoebel for
invaluable alpha testing and Paul-Michael Agapow for providing the
*multilocus* C++ source code for reference. We thank Sydney Everhart and
Brian Knaus for comments that significantly improved this manuscript.
Mention of trade names or commercial products in this manuscript are
solely for the purpose of providing specific information and do not
imply recommendation or endorsement.


  **Method/Index**                                       **Citation**                             **Function(s) in *poppr***
  ---------------------------- ---------------------------------------------------------------- ----------------------------
  Expected MLG (rarefaction)      @hurlbert1971nonconcept, @heck1975explicit (for std. err.)                       `poppr()`
  $H$                                              @shannon2001mathematical                                        `poppr()`
  $G$                                               @stoddart1988genotypic                                         `poppr()`
  $H_{exp}$                                               @Nei:1978                                                `poppr()`
  $E_{5}$                       [@Grunwald:2006; @pielou1975ecological; @ludwig1988statistical]                     `poppr()`
  $I_A$/$\bar{r}_d$             [@Brown:1980; @Smith:1993] ($I_A$), @Agapow:2001 ($\bar{r}_d$)                         `ia()`
                                                                                                                   `poppr()`
  Clone correction               [@milgroom1996recombination; @Grunwald:2003; @Grunwald:2006]                `clonecorrect()`
                                                                                                                   `poppr()`
  Minimum Spanning Networks                                @igraph                                             `poppr.msn()`
                                                                                                               `bruvo.msn()`
  Bruvo's Distance                                       @Bruvo:2004                                          `bruvo.dist()`
                                                                                                               `bruvo.msn()`
                                                                                                              `bruvo.boot()`
  Bootstrapping                                              @ape                                             `bruvo.boot()`
  Neighbor Joining                                           @ape                                             `bruvo.boot()`
  UPGMA                                                   @phangorn                                           `bruvo.boot()`



                  **Haploids**   **Diploids**   $\bar{r}_d$   **All Platforms**   **Batch Analysis**
  -------------- -------------- -------------- ------------- ------------------- --------------------
  *poppr*             Yes            Yes            Yes              Yes                 Yes
  *LIAN*              Yes             No            No               Yes                 Yes
  *multilocus*        Yes            Yes            Yes              No                   No



                  **$I_A$ (seconds)**   **Bruvo's distance (seconds)**
  -------------- --------------------- --------------------------------
  *poppr*                13.4                        0.3
  *polysat*                -                         58.3
  *multilocus*           547.2                        -

![\[fig:FIGURE\_1\] Distribution of 12 multilocus genotypes from the
`Finland` population of the `H3N2` SNP data set
[@Jombart:2008].](figures/Finland/Finland.png){width="0.7\columnwidth"}

![\[fig:FIGURE\_2\] Visualizations of tests for linkage disequilibrium,
where observed values (blue dashed lines) of $I_A$ and $\bar{r}_d$ are
compared to histograms showing results of 999 permutations using method
1 in table \[table1\]. Results are shown for the sexual population `5`
of the `nancycats` data set \[@Jombart:2008\] (a) and for the clonal
`Athena` population of the `Aeut` data set \[@Grunwald:2003\] (b)
.](figures/nancy_5/nancy_5.png){width="0.7\columnwidth"}

![\[fig:FIGURE\_3\] Example minimum spanning network using Bruvo's
distance on a simulated partially clonal data set with 50 individuals
genotyped over 10 microsatellite loci. Each node represents a unique
multilocus genotype. Node shading (colors) represent population
membership, while edge widths and shading represent relatedness. Edge
length is
arbitrary.](figures/bruvo_color/bruvo_color.png){width="0.7\columnwidth"}

![\[fig:FIGURE\_4\] UPGMA tree produced from Bruvo's distance with 1000
bootstrap replicates (node values greater than 50% are shown). Data from
population `9` of the `nancycats` data set
[@Jombart:2008].](figures/nancy9_tree/nancy9-boot-redo.png){width="0.7\columnwidth"}
