
```{r, results = "asis", echo = FALSE}
cat(beaverdown::iflatex("\\singlespacing"))
```

# *Poppr*: an R Package For Genetic Analysis of Populations With Clonal, Partially Clonal, and/or Sexual Reproduction

```{r, results = "asis", echo = FALSE}
cat(beaverdown::iflatex("\\begin{center}"))
```

Zhian N. Kamvar, Javier F. Tabima, and Niklaus J. Grünwald


```{r, results = "asis", echo = FALSE}
out <- "
\\end{center}
\\vspace*{\\fill}
"
cat(beaverdown::iflatex(out))
```
Journal: **PeerJ**

PO Box 614, Corte Madera, CA 94976, USA    

Published 2014-03-04, Issue: **2**:e281, 
DOI: [10.7717/peerj.281](https://dx.doi.org/10.7717/peerj.281)

```{r, results = "asis", echo = FALSE}
out <- "
\\doublespacing
\\newpage
"
cat(beaverdown::iflatex(out))
```

## Abstract


Many microbial, fungal, or oomcyete populations violate assumptions for
population genetic analysis because these populations are clonal, admixed,
partially clonal, and/or sexual. Furthermore, few tools exist that are
specifically designed for analyzing data from clonal populations, making
analysis difficult and haphazard. We developed the R package poppr providing
unique tools for analysis of data from admixed, clonal, mixed, and/or sexual
populations. Currently, poppr can be used for dominant/codominant and
haploid/diploid genetic data. Data can be imported from several formats
including GenAlEx formatted text files and can be analyzed on a user-defined
hierarchy that includes unlimited levels of subpopulation structure and clone
censoring. New functions include calculation of Bruvo's distance for
microsatellites, batch-analysis of the index of association with several indices
of genotypic diversity, and graphing including dendrograms with bootstrap
support and minimum spanning networks. While functions for genotypic diversity
and clone censoring are specific for clonal populations, several functions found
in poppr are also valuable to analysis of any populations. A manual with
documentation and examples is provided. Poppr is open source and major releases
are available on CRAN: http://cran.r-project.org/package=poppr. More supporting
documentation and tutorials can be found under 'resources' at:
http://grunwaldlab.cgrb.oregonstate.edu/.