# Spatial and Temporal Analysis of Populations of the Sudden Oak Death Pathogen in Oregon Forests

Zhian N. Kamvar, Meredeth M. Larsen, Alan M. Kanaskie, Everett M.
Hansen, and Niklaus J. Grünwald


```{r, results = "asis", echo = FALSE}
cat(beaverdown::iflatex("\\vspace*{\\fill}"))
```


**Published in Phytopathology 2015-07**        

DOI: [10.1094/PHYTO-12-14-0350-FI](http://dx.doi.org/10.1094/PHYTO-12-14-0350-FI)

```{r, results = "asis", echo = FALSE}
cat(beaverdown::iflatex("\\newpage"))
```


## Abstract

Sudden oak death caused by the oomycete *Phytophthora ramorum* was first
discovered in California toward the end of the 20th century and subsequently
emerged on tanoak forests in Oregon before its first detection in 2001 by aerial
surveys. The Oregon Department of Forestry has since monitored the epidemic and
sampled symptomatic tanoak trees from 2001 to the present. Populations sampled
over this period were genotyped using microsatellites and studied to infer the
population genetic history. To date, only the NA1 clonal lineage is established
in this region, although three lineages exist on the North American west coast.
The original introduction into the Joe Hall area eventually spread to several
regions: mostly north but also east and southwest. A new introduction into
Hunter Creek appears to correspond to a second introduction not clustering with
the early introduction. Our data are best explained by both introductions
originating from nursery populations in California or Oregon and resulting from
two distinct introduction events. Continued vigilance and eradication of nursery
populations of *P. ramorum* are important to avoid further emergence and
potential introduction of other clonal lineages.